Connecting every bit of knowledge: \\ The Structure of Wikipedia's first link network

### 2016-02-28 writeup/thesis format/thesis.tex
edits post peter comments

\documentclass[pre,twocolumn,twoside,superscriptaddress,floatfix, aps, 10pt]{revtex4-1}

\usepackage{currfile}
\lefthyphenmin=3
\righthyphenmin=2

\usepackage{graphicx,epsfig,verbatim,enumerate}
\usepackage{amssymb,amsmath}
\usepackage{ifthen}

% Fonts and typesetting settings
\usepackage[sc]{mathpazo}
\usepackage[T1]{fontenc}
\linespread{1.05} % Palatino needs more space between lines
\usepackage{microtype}
\usepackage{textcomp}

\usepackage{longtable}

\usepackage{mathtools}

\newboolean{twocolswitch}

\newcommand{\www}[1]{\url{#1}}
\newcommand{\req}[1]{(\ref{#1})}
\newcommand{\Req}[1]{Eq.~(\ref{#1})}

% Lettrines
\usepackage{lettrine}

%useful shortcuts
\def\R{\ensuremath{\mathbb{R}}} %\ensuremath adds math mode, if forgotten
\def\Q{\ensuremath{\mathbb{Q}}}
\def\N{\ensuremath{\mathbb{N}}}
\def\Z{\ensuremath{\mathbb{Z}}}
\def\C{\ensuremath{\mathbb{C}}}

%shorcuts with arguments
\newcommand{\abs}[1]{\left\vert#1\right\vert} %nice absolute values
\newcommand{\bt}[1]{\textbf{#1}} %bold
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}} %aligned equations
\newcommand{\norm}[1]{\left\lVert#1\right\rVert} %vector norm
\renewcommand{\eq}[1]{\begin{align*}#1\end{align*}} %aligned equations

%piecewise function

%\begin{displaymath}
%   f(x) = \left\{
%     \begin{array}{lr}
%       1 & : x \in \mathbb{Q}\\
%       0 & : x \notin \mathbb{Q}
%     \end{array}
%   \right.
%\end{displaymath}

%environment
\newcommand{\tab}{\phantom{ssss}}

\usepackage{hyperref}
\usepackage{color}
\newcommand{\todo}[1]{\noindent\textcolor{blue}{{$\Box$ #1}}}

%colors
\definecolor{javagreen}{rgb}{0.25,0.5,0.35} %dark green color
\definecolor{lightblue}{rgb}{0.149,0.545,0.824} %solarized blue
\definecolor{sred}{rgb}{0.863, 0.196, 0.184} %solarized red

\newcommand{\blue}[1]{{\leavevmode\color{lightblue}{#1}}} %solarized blue 
\newcommand{\green}[1]{{\leavevmode\color{javagreen}{#1}}} %command for green
\newcommand{\red}[1]{{\leavevmode\color{sred}{#1}}} %solarized red
\newcommand{\gray}[1]{{\leavevmode\color[gray]{0.5}{#1}}} %gray text
\newcommand{\Prob}[1]{{\rm Pr}\left(#1\right)}

\setboolean{twocolswitch}{true}

@@

\title{\protect
Connecting every bit of knowledge: \\
The Structure of Wikipedia's first link network
}

\date{\today}

\author{
\firstname{Mark}
\surname{Ibrahim}
}
\email{mark.s.ibrahim@uvm.edu}

\affiliation{Department of Mathematics \& Statistics, 
    Computational Story Lab,  \\
    Vermont Complex Systems Center,
    Vermont Advanced Computing Core,
    The University of Vermont, Burlington, VT 05401.}

\author{
\firstname{Christopher}
\surname{M. Danforth}
}
\email{chris.danforth@uvm.edu}

\affiliation{Department of Mathematics \& Statistics, 
    Computational Story Lab,  \\
    Vermont Complex Systems Center,
    Vermont Advanced Computing Core,
    The University of Vermont, Burlington, VT 05401.}

\author{
\firstname{Peter}
\surname{Sheridan Dodds}
}

\email{peter.dodds@uvm.edu}

\affiliation{Department of Mathematics \& Statistics, 
    Computational Story Lab,  \\
    Vermont Complex Systems Center,
    Vermont Advanced Computing Core,
    The University of Vermont, Burlington, VT 05401.}

\begin{abstract}
  \protect
Apples, porcupines, and the most obscure Dylan song---is every topic a few clicks from Philosophy? 
Within Wikipedia, the surprising answer is yes: nearly all 
paths lead to Philosophy.
Wikipedia is the largest, most meticulously indexed collection of human knowledge ever amassed. 
More than information about a topic, Wikipedia is a web of naturally emerging relationships.  
By following the first link in an article, we connect entries to form a directed network: Wikipedia's First Link Network. 
Here, we study the English edition of Wikipedia's First Link Network for insight into how the many topics, ideas, people, objects, and events
are related and organized.  
We algorithmically parse all 4.7 million articles to construct a map of Wikipedia's First Link Network. 
We traverse every possible path through the network, 
measuring the accumulation of first links, path lengths, basins, cycles, and the influence a particular article exerts in shaping the 
network.
We discover many scale-free distributions describing path length, 
accumulation, and influence; find Philosophy at a salient center; and uncover a flow from specific to general 
culminating around fundamental notions such as Community, State, and Science. 
Curiously, we also observe a gravitation towards topical articles including Health Care and Fossil Fuel.
These findings enrich our view of the connections and structure of
Wikipedia's ever growing store of knowledge.

@@

\maketitle

@@

The authors of a Wikipedia article choose where and whether to include a 
reference to another Wikipedia article
in the HTML markup.
For example, the authors of 
\href{https://en.wikipedia.org/wiki/Train}{``Train''}
chose
``Amtrak's Acela Express,'' ``steam,'' and ``head-end power'' 
among others as relevant articles to reference in describing ``Train''
\cite{wiki_train}
.
By focusing our attention on the main body of an article---excluding
elements in the side bars and headings---we capture the introduction to the topic.
The first link marks the earliest moment in a topic's introduction where the 
authors choose to directly reference another article. 
While the many links in an article may reference relevant details to a topic, 
the first link is an association within the introductory 
description. While 
``Amtrak's Acela Express,'' ``steam,'' and ``head-end power'' 
are links detailing particulars, the first link, ``rail transport,''
is the topic the authors associate with ``Train'' in the introduction. 
Collectively, the first links provide a pragmatic and interpretable 
means to connect each article to another.

By following the first hyperlink to another article in the English edition of
Wikipedia as of November, 2014, we connect the ideas in one article to those in another, ultimately forming a directed network: 
{\it Wikipedia's First Link Network} (FLN).

@@

What information could the First Links possibly reveal?
Inspired by the claim that the majority of First Links lead to the 
``Philosophy'' article---popularized by an 
\href{https://xkcd.com/903/}{xkcd}
comic and subsequently
discussed in blog posts 
\cite{mat_blog, Ilmari_first_links, xkcd}
---we holistically study 
Wikipedia's FLN as a map relating areas of human knowledge. 
The map is a hierarchical taxonomy where, for example, ``Train'' links to a parent node, 
``Rail Transport,'' while many child nodes, ``Steel'' and ``Horsepower,'' link to 
``Train.'' Unlike previous taxonomies
created by individuals
\cite{locke, descartes, aristotle}
or a select group of individuals 
\cite{hist_thesaurus}, 
the organization of ideas in the FLN 
does not emerge from a centralized effort. 
Rather each link in the FLN emerges from the choice of each author and ensuing edits.
Collectively, the FLN is a wealth of relations among inventions, places,
figures, objects, and events across space and time.

@@

Our greater interest is lies in the richer dynamics of how many articles fit together---how links create a flow through the FLN.
We construct a path through the FLN with each article 
leading to the next through the first link. 
For example ``Train'' has a first link to ``Conveyance of Passengers and Goods,'' which is itself
an article with a first link to ``Goods'' and so on. 
As shown in Fig.~\ref{fig:Train First Links}, the path starting at ``Train'' 
contains ``Goods,'' ``Economics,'' ``Social Science,''
leading ultimately to ``Philosophy''.

@@

Starting at each article, we construct a path through the FLN, 
to map the flow of connections among articles.
The method is order agnostic with respect to which articles are selected first. As long as each article is selected eventually, the resulting metrics are equivalent.

We use the notion of a path through the FLN to describe 
the structure of the FLN as a flow relating topics. 
Grounded in the dynamics of river networks
\cite{geo_basins},
we characterize features of the FLN as a flow through first links.
Previous studies have used flow to characterize the structure of river networks
\cite{dodds} and describe the organization of food systems through transportation networks
\cite{food_webs}.
With paths to mark flow through the FLN, we develop metrics to 
gauge the accumulation of references, 
the length of the path relating articles, and the influence a particular
article exerts in shaping the flow of links through the FLN.

@@

       \bibitem{coverage} Halavais, Alexander, and Derek Lackaff. ``An analysis of topical coverage of Wikipedia.'' Journal of Computer‐Mediated Communication 13.2 (2008): 429-440.

@@

    \bibitem{locke} Bolton, Martha Brandt. ``The Taxonomy of Ideas in Locke’s {\it Essay},'' The Cambridge Companion to Locke's 'Essay Concerning Human Understanding'. Ed.Lex Newman.. 1st ed. Cambridge: Cambridge University Press, 2007. 67-100. Cambridge Companions Online. Web. 13 November 2015. http://dx.doi.org/10.1017/CCOL0521834333.004.

>>>>

\documentclass[11pt]{report}

\usepackage[latin2]{inputenc}
\usepackage{amsmath}
\usepackage[pdftex]{graphicx}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage[left=1.5in]{geometry}
\usepackage{geometry}
\pagestyle{fancyplain}
\usepackage{wrapfig}
\usepackage{setspace}
\usepackage{epsfig}
\usepackage{color}
\usepackage{tikz}
\usepackage{thesis_MTG_3}
\usepackage{cancel}
\usepackage{cite}

% ==================================
% Set up title page
% ==================================
\title{
Connecting every bit of knowledge: \\
The Structure of Wikipedia's first link network
}
\author{Mark Ibrahim}
\defensedate{March, 2016}
\thesis
\masterscience
\matsci
\maygrad
\advisor{Christopher M. Danforth, Ph.D.} %advisor
\readerone{Christopher M. Danforth, Ph.D.}
\readertwo{Peter Sheridan Dodds, Ph.D.}
\chair{Randall Harp, Ph.D.}

@@

\maketitle
\makeacceptance
\pagenumbering{roman}

\begin{abstract}
Apples, porcupines, and the most obscure Dylan song---is every topic a few clicks from Philosophy? 
Within Wikipedia, the surprising answer is yes: nearly all 
paths lead to Philosophy.
Wikipedia is the largest, most meticulously indexed collection of human knowledge ever amassed. 
More than information about a topic, Wikipedia is a web of naturally emerging relationships.  
By following the first link in an article, we connect entries to form a directed network: Wikipedia's First Link Network. 
Here, we study the English edition of Wikipedia's First Link Network for insight into how the many topics, ideas, people, objects, and events
are related and organized.  
We algorithmically parse all 4.7 million articles to construct a map of Wikipedia's First Link Network. 
We traverse every possible path through the network, 
measuring the accumulation of first links, path lengths, basins, cycles, and the influence a particular article exerts in shaping the 
network.
We discover many scale-free distributions describing path length, 
accumulation, and influence; find Philosophy at a salient center; and uncover a flow from specific to general 
culminating around fundamental notions such as Community, State, and Science. 
Curiously, we also observe a gravitation towards topical articles including Health Care and Fossil Fuel.
These findings enrich our view of the connections and structure of
Wikipedia's ever growing store of knowledge.

@@

\begin{acknowledgements}
    Thank you to my advisors \dots
\end{acknowledgements}

\tableofcontents
\newpage

\listoffigures
\newpage

\doublespacing

@@

The authors of a Wikipedia article choose where and whether to include a 
reference to another Wikipedia article
in the HTML markup.
For example, the authors of 
\href{https://en.wikipedia.org/wiki/Train}{``Train''}
chose
``Amtrak's Acela Express,'' ``steam,'' and ``head-end power'' 
among others as relevant articles to reference in describing ``Train''
\cite{wiki_train}
.
By focusing our attention on the main body of an article---excluding
elements in the side bars and headings---we capture the introduction to the topic.
The first link marks the earliest moment in a topic's introduction where the 
authors choose to directly reference another article. 
While the many links in an article reference relevant details, 
the first link is an association within a topic's introductory 
description. While 
``Amtrak's Acela Express,'' ``steam,'' and ``head-end power'' 
are links detailing particulars, the first link, ``rail transport,''
is the topic the authors associate with ``Train'' in the introduction. 
[TODO:add a few more examples]
Collectively, the first links provide a pragmatic and interpretable 
means to connect each article to another.

By following the first hyperlink to another article in the English edition of
Wikipedia as of November, 2014, we connect the topic of one article to that of another, ultimately forming a directed network: 
{\it Wikipedia's First Link Network} (FLN).

@@

The FLN is a wealth of relations among inventions, places,
figures, objects, and events across space and time.
``Train'' for example, links to a parent node, ``Rail Transport,''
while many child nodes such as ``Steel'' and ``Horsepower'' link to ``Train.''
Unlike previous taxonomies created by individuals
\cite{locke, descartes, aristotle, hist_thesaurus},
the relations in the FLN emerge without a centralized effort 
as the aggregate of each article's authors choice of first link.

@@

Our greater interest is lies in the richer dynamics of how many articles fit together.
Inspired by the claim that the majority of First Links lead to
``Philosophy''---popularized by an 
\href{https://xkcd.com/903/}{xkcd}
comic and subsequently
discussed in blog posts 
\cite{mat_blog, Ilmari_first_links, xkcd}
---
we holistically study how the FLN  yields a flow of connections.
We construct a path through the FLN with each article 
leading to the next through the first link. 
For example ``Train'' has a first link to ``Conveyance of Passengers and Goods,'' which is itself
an article with a first link to ``Goods'' and so on. 
As shown in Fig.~\ref{fig:Train First Links}, the path starting at ``Train'' 
contains ``Goods,'' ``Economics,'' ``Social Science,''
leading ultimately to ``Philosophy''.

@@

Starting at each article, we construct a path through the FLN, 
to map the flow of connections among articles.
The method is order agnostic with respect to which articles are selected first. As long as each article is selected eventually, the resulting metrics are equivalent.
Grounded in the dynamics of river networks
\cite{geo_basins},
we characterize features of the FLN as a flow through first links.
Previous studies have used flow to characterize the structure of river networks
\cite{dodds} and describe the organization of food systems through transportation networks
\cite{food_webs}.
With paths to mark flow through the FLN, we develop metrics to 
gauge the accumulation of references, 
the length of the path relating articles, and the influence a particular
article exerts in shaping the flow of links through the FLN.

@@

       \bibitem{coverage} Halavais, Alexander, and Derek Lackaff. ``An analysis of topical coverage of Wikipedia.'' Journal of Computer Mediated Communication 13.2 (2008): 429-440.

@@

    \bibitem{locke} Bolton, Martha Brandt. ``The Taxonomy of Ideas in Locke\textquotesingle s {\it Essay},'' The Cambridge Companion to Locke's ``Essay Concerning Human Understanding.'' Ed.Lex Newman.. 1st ed. Cambridge: Cambridge University Press, 2007. 67-100. Cambridge Companions Online. Web. 13 November 2015. http://dx.doi.org/10.1017/CCOL0521834333.004.
